---
title: "Solutions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Solutions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Suggested answers to the workshop questions are below. You might have some different code e.g. to customise the volcano plot as you like. Feel free to comment on any of these solutions in the workshop website as described [here](https://github.com/stemangiola/rpharma2020_tidytranscriptomics/blob/master/CONTRIBUTING.md).

```{r out.width = "70%", message=FALSE, warning=FALSE}
# load libraries

# tidyverse core packages
library(tibble)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(purrr)
library(ggplot2)

# tidyverse-friendly packages
library(plotly)
library(ggrepel)
library(tidyHeatmap)
library(tidybulk)

# load data
data("pasilla", package = "rpharma2020tidytranscriptomics")

# create tidybulk tibble
counts_tt <-
    pasilla %>%
    tidybulk() %>%
    mutate(symbol = AnnotationDbi::mapIds(org.Dm.eg.db::org.Dm.eg.db, keys=as.character(feature), keytype = "FLYBASE", column="SYMBOL", multiVals = "first"))

# filter counts
counts_filtered <- counts_tt %>% keep_abundant(factor_of_interest = condition)

# scale counts
counts_scaled <- counts_filtered %>% scale_abundance()

# create density plots
counts_scaled %>%
	pivot_longer(cols = c("counts", "counts_scaled"), names_to = "source", values_to = "abundance") %>%
  ggplot(aes(x=abundance + 1, group=sample, color=condition)) +
	geom_density() +
	facet_wrap(~source) +
	scale_x_log10() +
	theme_bw()
```

# Question 1.1

What is the Fraction of Variance for PC1 and PC2? 

```{r}
counts_scal_PCA <-
  counts_scaled %>%
  reduce_dimensions(method="PCA")
```


What do PC1 and PC2 represent?

```{r out.width = "70%"}
counts_scal_PCA %>%
	pivot_sample() %>%
	ggplot(aes(x=PC1, y=PC2, colour=condition, shape=type)) + 
  geom_point() +
  geom_text_repel(aes(label=sample), show.legend = FALSE) +
	theme_bw()
```

Answer: PC1 represents variance due to treatment effect(treated vs untreated). PC2 represents variance due to sequencing type single vs paired.


```{r}
counts_de <-
  counts_tt %>%
  test_differential_abundance(.formula = ~ 0 + condition + type, 
                              .contrasts = c("conditiontreated - conditionuntreated"), 
                              omit_contrast_in_colnames = TRUE)
```

# Question 1.2 

How many DE genes are there for treated vs untreated (FDR < 0.05)?

```{r}
counts_de %>% 
  filter(FDR < 0.05) %>% 
  summarise(num_de = n_distinct(feature))
```


# Question 1.3

What is the FBgn id of the 10th most differentially expressed gene (by smallest P value)?

```{r}
topgenes <- counts_de %>%
	pivot_transcript() %>%
  arrange(PValue) %>%
  head(10)

topgenes
```	


# Question 1.4

What code can generate a heatmap of variable genes (starting from count_scaled)?  

```{r out.width = "70%"}
counts_scaled %>% 
	
	# extract 500 most variable genes
	keep_variable( .abundance = counts_scaled, top = 500) %>%
	
	# create heatmap
	heatmap(
	      .column = sample,
	      .row = feature,
	      .value = counts_scaled,
	      annotation = c(condition, type),
	      transform = log1p 
	  )
```

# Question 1.5

What code can you use to visualise expression of the pasilla gene (gene id: FBgn0261552) 

```{r out.width = "70%"}
counts_scaled %>%
	
	# extract counts for pasilla gene
	filter(feature == "FBgn0261552") %>%
	
	# make stripchart
	ggplot(aes(x = condition, y = counts_scaled + 1, fill =condition, label = sample)) +
	geom_boxplot() +
	geom_jitter() +
	scale_y_log10()+
	theme_bw()
```

# Question 1.6

What code can generate an interactive volcano plot that has gene ids showing on hover?  

```{r out.width = "70%"}
p <- counts_de %>%
	pivot_transcript() %>%

  # Subset data
	mutate(significant = FDR<0.05 & abs(logFC) >=2) %>%

  # Plot
	ggplot(aes(x = logFC, y = PValue, label=feature)) +
	geom_point(aes(color = significant, size = significant, alpha=significant)) +
	geom_text_repel() +
	
	# Custom scales
	scale_y_continuous(trans = "log10_reverse") +
	scale_color_manual(values=c("black", "#e11f28")) +
	scale_size_discrete(range = c(0, 2)) +
	theme_bw()

ggplotly(p, tooltip = c("text"))
```
Tip: You can use "text" instead of "label" if you don't want the column name to show up in the hover e.g. above will give "FBgn0261552" rather than "feature:FBgn0261552".

# Question 1.7

What code can generate a heatmap of the top 100 DE genes?

```{r out.width = "70%"}
top100 <- 
	counts_de %>%
	pivot_transcript() %>%
	arrange(PValue) %>%
	head(100)

counts_scaled %>% 
  filter(feature %in% top100$feature) %>%
	heatmap(
	      .column = sample,
	      .row = feature,
	      .value = counts_scaled,
	      annotation = c(condition, type),
	      transform = log1p 
	  )
```

# Question 2.1

How can I compare the amount of differential abundant transcripts for the different methods?

```{r}
# Set up data
pasilla_de <- 
  rpharma2020tidytranscriptomics::pasilla %>% 
  
  # Convert SE object to tibble
  tidybulk %>%
  
   # Scale abundance for plotting
  identify_abundant(factor_of_interest=condition) 

de_all <- 
  
  pasilla_de %>%
  
  # edgeR QLT
  test_differential_abundance(
    ~ condition + type, 
    method = "edger_quasi_likelihood",
    prefix = "edgerQLT_"
  )  %>%
  
  # edgeR LRT
  test_differential_abundance(
    ~ condition + type, 
    method = "edger_likelihood_ratio",
    prefix = "edgerLR_"
  )  %>%
  
  # limma-voom
  test_differential_abundance(
    ~ condition + type, 
    method = "limma_voom",
    prefix = "voom_"
  ) %>%
  
  # DESeq2
  test_differential_abundance(
    ~ condition + type, 
    method = "deseq2",
    prefix = "deseq2_"
  ) 
```

```{r out.width = "70%"}
de_all %>%
	
	# Subset transcript information
	pivot_transcript() %>%
	
	# Reshape for nesting
	pivot_longer(
		cols = -c(feature, .abundant),
		names_sep = "_", 
		names_to = c("method", "statistic"), 
		values_to = "value"
	) %>%
	
	# Filter statistic
	filter(statistic %in% c("FDR", "adj.P.Val", "padj")) %>%
	filter(value < 0.05) %>%
	
	# Nesting
	dplyr::count(method)
	
```

# Question 2.2 

How can I trace the adjusted p-value rank for methods?

```{r out.width = "70%"}
# significance trend
trend_plot = 
	de_all %>%
	
	# Subset transcript information
	pivot_transcript() %>%
	
	# Reshape for nesting
	pivot_longer(
		cols = -c(feature, .abundant),
		names_sep = "_", 
		names_to = c("method", "statistic"), 
		values_to = "value"
	) %>%
	
	# Filter statictic
	filter(statistic %in% c("FDR", "adj.P.Val", "padj")) %>%
	filter(value %>% is.na %>% `!`) %>%

	# Nesting
	nest(data = -method) %>%
	mutate(data = map(
		data,
		~ .x %>% 
			arrange(value) %>%
			mutate(rank = 1:n())
	)) %>%
	unnest(data) %>%
	
	ggplot(aes(rank, value, color = method)) + 
	geom_line() 

trend_plot

# Zooming on highly significant events
trend_plot + 
	scale_x_log10() + 
	scale_y_log10()
	
```
## What is the most abundant cell type overall in BRCA samples?

```{r eval=FALSE}
BRCA_cell_type_long %>% 
	group_by(cell_type) %>% 
	summarise(m = median(proportion)) %>% 
	dplyr::arrange(dplyr::desc(m))

```



