---
title: "Supplementary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Supplementary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, warning=FALSE}
# load libraries
library(airway)
library(tibble)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(purrr)
library(ggplot2)
library(ggrepel)
library(tidyHeatmap)
library(tidybulk)
```

Plot settings. Set the colours and theme we will use for our plots.

```{r}
# Use colourblind-friendly colours
friendly_cols <- dittoSeq::dittoColors()

# Set theme
custom_theme <-
  list(
    scale_fill_manual(values = friendly_cols),
    scale_color_manual(values = friendly_cols),
    theme_bw() +
      theme(
        panel.border = element_blank(),
        axis.line = element_line(),
        panel.grid.major = element_line(size = 0.2),
        panel.grid.minor = element_line(size = 0.1),
        text = element_text(size = 12),
        legend.position = "bottom",
        #aspect.ratio = 1,
        strip.background = element_blank(),
        axis.title.x = element_text(margin = margin(t = 10, r = 10, b = 10, l = 10)),
        axis.title.y = element_text(margin = margin(t = 10, r = 10, b = 10, l = 10)),
        axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)
      )
  )
```

# Part 1 Bulk RNA-seq Core

## How to start from tables

```{r}
# create some example tables to use
data(airway)

# counts table
counts <- assay(airway) %>%
  as_tibble(rownames = "geneID")

# sample information table
sampleinfo <- colData(airway) %>%
  as_tibble(rownames = "sample")

# data preprocessing

counts_tt <- 
	# convert to tidy format
	pivot_longer(counts, cols = starts_with("SRR"), names_to = "sample", values_to = "counts") %>%

	# get gene symbols
	ensembl_to_symbol(geneID) %>%

	# order the columns for tidybulk
	select(sample, geneID, counts, transcript) %>%

	# add the sample info
	left_join(sampleinfo) %>%

	# shorten sample name
	mutate(sample=str_remove(sample, "SRR1039")) %>%

	# convert to tidybulk tibble
	tidybulk(.sample=sample, .transcript=geneID, .abundance=counts)
```

## How to count reads per sample

```{r}
counts_tt %>% 
	group_by(sample) %>% 
	summarise(total_reads=sum(counts))
```

We can also check how many counts we have for each sample by making a bar plot. This helps us see whether there are any major discrepancies between the samples more easily.

```{r out.width = "70%"}
ggplot(counts_tt, aes(x=sample, weight=counts, fill=sample)) + 
	geom_bar() +
	theme_bw()
```

As we are using ggplot2, we can also easily view by any other variable that's a column in our dataset, such as cell line, simply by changing `fill`.

We can colour by dex treatment.

```{r out.width = "70%"}
ggplot(counts_tt, aes(x=sample, weight=counts, fill=dex)) + 
	geom_bar() +
	theme_bw()
```
We can colour by cell line.

```{r out.width = "70%"}
ggplot(counts_tt, aes(x=sample, weight=counts, fill=cell)) + 
	geom_bar() +
	theme_bw()
```


## How to examine normalised counts with boxplots

```{r out.width = "70%"}
# filter counts
counts_filtered <- counts_tt %>% keep_abundant(factor_of_interest = dex)

# scale counts
counts_scaled <- counts_filtered %>% scale_abundance()

# create box plots
counts_scaled %>%
	pivot_longer(cols = c("counts", "counts_scaled"), names_to = "source", values_to = "abundance") %>%
	ggplot(aes(x=sample, y=abundance + 1, fill=dex)) +
	geom_boxplot() +
	geom_hline(aes(yintercept = median(abundance + 1)), colour="red") +
	facet_wrap(~source) +
	scale_y_log10() +
	theme_bw()
```

## How to create MDS plot

```{r out.width = "70%"}
airway %>%
	tidybulk() %>%
    keep_abundant(factor_of_interest = dex) %>%
	scale_abundance() %>%
	reduce_dimensions(method="MDS", scale = FALSE) %>%
	pivot_sample() %>%
	ggplot(aes(Dim1, Dim2, color = dex)) +
	geom_point()
```


## How to create MA plot

MA plots enable us to visualise amount of expression (logCPM) versus logFC. Highly expressed genes are towards the right of the plot. We can also colour significant genes (e.g. genes with FDR < 0.05) 

```{r out.width = "70%"}
# perform differential testing
counts_de <- 
	counts_tt %>%
	test_differential_abundance(
		.formula = ~ 0 + dex + cell,
		.contrasts = c("dextrt - dexuntrt"),
		omit_contrast_in_colnames = TRUE
	)

# maplot, minimal
counts_de %>%
	pivot_transcript() %>%
	ggplot(aes(x=logCPM, y=-logFC, colour=FDR < 0.05)) +
	geom_point()+
	theme_bw()
```

A more informative MA plot, integrating some of the packages in tidyverse.

```{r out.width = "70%", warning=FALSE}
counts_de %>%
	pivot_transcript() %>%
	
	# Subset data
	mutate(significant = FDR<0.05 & abs(logFC) >=2) %>%
	mutate(transcript = ifelse(abs(logFC) >=5, as.character(transcript), "")) %>%
	
	# Plot
	ggplot(aes(x = logCPM, y = logFC, label=transcript)) +
	geom_point(aes(color = significant, size = significant, alpha=significant)) +
	geom_text_repel() +
	scale_color_manual(values=c("black", "#e11f28")) +
	scale_size_discrete(range = c(0, 2)) +
	theme_bw()
```


## How to perform gene enrichment analysis

To run below you'll need the `clusterProfiler` and `org.Hs.eg.db` packages. This is just one suggestion, adapted from [here](https://simon-anders.github.io/data_analysis_course/lecture9.html). If you have other suggestions for how to do a 'tidy' pathway analysis feel free to [let us know](https://github.com/stemangiola/rpharma2020_tidytranscriptomics/blob/master/CONTRIBUTING.md).

```{r eval=FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)

# extract all genes tested for DE
res <- counts_de %>% 
    pivot_transcript()

# GO terms
egoCC <- res %>%
    filter(FDR < 0.1 & logFC > 0 ) %>%
    pull( "transcript" ) %>%
    enrichGO(
      OrgDb = org.Hs.eg.db,
      keyType = 'SYMBOL',
      ont = "BP",
      universe = (res %>% pull( "transcript" ) ) )

dotplot(egoCC)
goplot(egoCC)
emapplot(egoCC)


# MSigDB Hallmark
gmtH <- read.gmt( "https://data.broadinstitute.org/gsea-msigdb/msigdb/release/6.2/h.all.v6.2.symbols.gmt" )
enrH <- enricher(
   gene = ( res %>% filter(FDR < 0.1 & logFC > 0) %>%
    pull( "transcript" ) ),
   TERM2GENE = gmtH,
   universe = ( res %>% pull( "transcript" ) ) )

dotplot( enrH )
emapplot(enrH)
```

# Part 2 Bulk RNA-seq Extended

## Nested analyses 

`tidybulk` allows for data nesting, using the `tidyr` utility `nest`. This is an extremely powerful tool as it enables easily performing analyses on data subsets.

## How to perform same analysis on subsets

Let's suppose we want to perform differential transcript abundance analysis independently for two different data subsets to compare results after the test


```{r}
pasilla_de <- 
  rpharma2020tidytranscriptomics::pasilla %>% 
  
  # Convert SummarizedExperiment object to tibble
  tidybulk %>%
  
  # Filter counts
  identify_abundant(factor_of_interest=condition) %>%
    
  # Scale abundance   
  scale_abundance() %>%
  
  # Nest
  nest(data = -type) %>%
  
  # Differential analysis
  mutate(data = map(
    data,
    ~  test_differential_abundance(.x, ~ condition) 
  )) %>%
  
  unnest(data)
```

Now we can for example compare the number of differentially transcribed genes and their co-expression

```{r}
pasilla_de %>%
  nest(data = -type) %>%
  mutate(
    number_of_differential = map_int(
      data, ~ .x %>% pivot_transcript() %>% filter(FDR < 0.05) %>% nrow
    ))
```

We can easily see which genes overlap, and plot them

```{r}
pasilla_de %>%
  filter(FDR < 0.05) %>%
  nest(data = -feature) %>%
  mutate(occurrences = map_int(data, ~ .x %>% distinct(type) %>% nrow)) %>%
  
  # We filter some of them
  filter(occurrences == 2) %>%
  dplyr::slice(1:6) %>%
  unnest(data) %>%
  
  # And plot
  ggplot(aes(type, counts_scaled +1, color=condition)) + 
  geom_point() + 
  facet_wrap(~feature) +
  scale_y_log10() +
  custom_theme
```

## How to perform analysis on subset and apply to full dataset

Let's suppose we want to identify the markers that distinguish epithelial from endothelial cells, and we also want to then visualise the abundance of those transcripts across many cell types to understand their cell type specificity.

```{r eval=FALSE}
cell_type_tt = rpharma2020tidytranscriptomics::cell_type_df %>% tidybulk(sample, symbol, count)
markers_df = 
  cell_type_tt %>%
  
  # Filter counts
  identify_abundant(factor_of_interest=cell_type) %>%
  
  # Scale abundance
  scale_abundance() %>%
  
  # Nest
  nest(data = everything()) %>%
  
  # Investigate one cell type pair
  mutate(comparison_data = map(
    data,
    ~ .x %>%
      filter(cell_type %in% c("endothelial", "epithelial")) %>%
      mutate(cell_type = as.character(cell_type) ) 
  )) %>%
  
  #test. We run on the two populations but we select data for all populations
  mutate(markers = map(
    comparison_data,
    ~ .x %>%
      
      # Differential transcription
      test_differential_abundance(
        ~ 0 + cell_type, 
        .contrasts = c("cell_typeendothelial - cell_typeepithelial"),
        action="only", 
        omit_contrast_in_colnames = TRUE
      ) %>%
      
      # Select markers
      filter(logFC > 2) %>% 
      dplyr::slice(1:10) %>% 
      pull(symbol)
    
  )) %>%
  
  # Add marker info to original data
  mutate(data = map2(data, markers, ~ .x %>% filter(symbol%in% .y))) %>%
  select(data) %>%
  unnest(data) 
```

Now we can see the abundance of the markers in all our cell types.

```{r eval=FALSE}
markers_df %>%
    ggplot(aes(cell_type, count_scaled +1 )) + 
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(size=0.3) +
    facet_wrap(~symbol, ncol=5) +
    coord_flip() +
    scale_y_log10() + 
    custom_theme
```
